---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---

# Project

Goal - To explore the NOAA Storm Database regarding severe weather events.

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The data for this assignment is Storm Data from the National Weather Service. Documentation for the data can be found at the [National Weather Service](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

## Data Processing

``` {r results='hide', message=FALSE, warning=FALSE}
### start necessary libraries
library(tidyverse)
library(dplyr)
library(lubridate)
library(rlist)
library(ggplot2)
library(knitr)

### set figure save path to figure/
knitr::opts_chunk$set(fig.path = 'figure/')

```

## Download data

Data comes in the form of a comma-separated-value file compressed via the bzip2 algorithm.

``` {r cache=TRUE}
### download data ZIP file if necessary, and extract if necessary
if (!file.exists('StormData.csv.bz2')) {
  if (download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2','StormData.csv.bz2')) {
    stop('Unable to download data file')
  }
}

StormData <- as.tibble(read.csv(bzfile('StormData.csv.bz2')))

```

Variables of interest

Variable    | Description                 
------------|-----------------------------
BGN_DATE    | date of event starting
BGN_TIME    | time of event starting
TIME_ZONE   | time zone of event
COUNTYNAME  | name of county
STATE       | state of the United STates
EVTYPE      | type of event
FATALITIES  | number of fatalities        
INJURIES    | number of injuries
PROPDMG     | value of property damage in dollars
PROPDMGEXP  | property damage dollar multiplier (K thousands, M millions, B billions)
CROPDMG     | value of crop damage in dollars
CROPDMGEXP  | crop damage dollar multiplier (K thousands, M millions, B billions)
REFNUM      | unique reference number for event


### Extract times of storm commencement

Extract time information from **BGN_DATE**, **BGN_TIME** and **TIME_ZONE**, and store in **BEGIN_TIME**.

``` {r}

### Information in BGN_TIME is not consistently stored
### some are in four-digit format e.g. '0000', '1300', '2350'. there is even one three-digit '000'
### others are in 11-character format e.g. '03:25:00 PM'
### convert all to 11-character format

StormData$BGN_calcTIME <- ifelse(str_length(StormData$BGN_TIME)<=4,
                             strftime(strptime(StormData$BGN_TIME,'%H%M'),'%I:%M:%S %p'),
                             as.character(StormData$BGN_TIME))
### more times which are in 'invalid' AM/PM format
### e.g. 15:30:00 PM, or 00:00:10 AM
### re-interpret these as 24-hour times
invalid.time.list <- is.na(strptime(StormData$BGN_calcTIME,'%I:%M:%S %p'))
StormData$BGN_calcTIME[invalid.time.list] <- 
  strftime(strptime(StormData[invalid.time.list,]$BGN_calcTIME,'%H:%M'), '%I:%M:%S %p')

### some times still remain invalid
print('Invalid times to be manually adjusted')
print(StormData[is.na(StormData$BGN_calcTIME),c('BGN_TIME','COUNTYNAME','STATE','REFNUM')])
### these times are 'impossible' times are 19:90 hr, or sometimes subsituted a '0 (zero)' for 'O (letter)'
### manually adjust these times, assuming simple transposition or substitution errors
StormData$BGN_calcTIME[StormData$REFNUM == 197399] <- '07:09:00 PM' ### was 1990
StormData$BGN_calcTIME[StormData$REFNUM == 223421] <- '12:01:00 AM' ### was 9999
StormData$BGN_calcTIME[StormData$REFNUM == 244548] <- '08:09:00 PM' ### was 2090
StormData$BGN_calcTIME[StormData$REFNUM == 247251] <- '03:08:00 PM' ### was 1580
StormData$BGN_calcTIME[StormData$REFNUM == 247767] <- '05:27:00 AM' ### was 0572
StormData$BGN_calcTIME[StormData$REFNUM == 248507] <- '01:00:00 PM' ### was 13O0

StormData$BEGIN_TIME <- as.POSIXct(strptime(paste(sub(' .*','',StormData$BGN_DATE),
                                       StormData$BGN_calcTIME),'%m/%d/%Y %I:%M:%S %p'))
```

**TIME_ZONE** information, in particular, requires tidying up and adjustment.

List all current timezones, and standardize capitalization

``` {r}
print(unique(StormData$TIME_ZONE))
```
``` {r}

### some timezone names use 'lower-case' characters, convert all to upper-case
StormData$TIME_ZONE <- as.factor(toupper(StormData$TIME_ZONE))

```

List records with dubious timezone names

```{r, echo=TRUE}
print(StormData[StormData$TIME_ZONE %in% c('UNK','CSC','ESY','SCT'), c('COUNTYNAME','STATE','TIME_ZONE','REFNUM')])
```

Fill in corrected timezones.

Convert 'daylight saving' timezones to 'standard' timezones, as Lubridate adjusts for daylight saving times automatically.

``` {r}
### replace missing timezones 'UNK'
### Oconee County, Georgia
StormData$TIME_ZONE[StormData$REFNUM == 23087] <- 'EST'
### Hawaii country, Hawaii
StormData$TIME_ZONE[StormData$REFNUM == 27702] <- 'HST'
### Coles county, Illinois
StormData$TIME_ZONE[StormData$REFNUM == 30669] <- 'CST'
### Rush County, Indiana
StormData$TIME_ZONE[StormData$REFNUM == 35244] <- 'EST'
### Hampden County, Massachusetts
StormData$TIME_ZONE[StormData$REFNUM == 67899] <- 'EST'
### Love County, Oklahoma
StormData$TIME_ZONE[StormData$REFNUM == 120704] <- 'CST'
### Perkins County, South Dakota
StormData$TIME_ZONE[StormData$REFNUM == 143198] <- 'MST'
### McLennan County, Texas
StormData$TIME_ZONE[StormData$REFNUM == 154637] <- 'CST'
### Armstrong, Texas
StormData$TIME_ZONE[StormData$REFNUM == 160236] <- 'CST'

### Chickasaw, Mississippi. likely mistaken timezone 'CSC'
StormData$TIME_ZONE[StormData$REFNUM == 200720] <- 'CST'
### Cambria, Pennsylvania, likely mistaken timezone 'ESY'
StormData$TIME_ZONE[StormData$REFNUM == 229824] <- 'EST'
### Shawano, Wisconsin, likely mistaken timezone 'SCT'
StormData$TIME_ZONE[StormData$REFNUM == 247612] <- 'CST'
### IAZ033, Indiana, likely mistaken timezone 'SCT'
StormData$TIME_ZONE[StormData$REFNUM == 201257] <- 'EST'

### convert daylight saving timezones to 'standard' time
StormData$TIME_ZONE[StormData$TIME_ZONE == 'ADT'] <- 'AST'
StormData$TIME_ZONE[StormData$TIME_ZONE == 'CDT'] <- 'CST'
StormData$TIME_ZONE[StormData$TIME_ZONE == 'EDT'] <- 'EST'
StormData$TIME_ZONE[StormData$TIME_ZONE == 'MDT'] <- 'MST'
StormData$TIME_ZONE[StormData$TIME_ZONE == 'PDT'] <- 'PST'

```

Adjust **BEGIN_TIME** according to **TIME_ZONE**.

``` {r}

### adjust time zones. roll=TRUE prevents 'NA' times as result of daylight savings
StormData[StormData$TIME_ZONE == 'AKS',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'AKS',]$BEGIN_TIME, tzone = 'America/Anchorage', roll = TRUE)
StormData[StormData$TIME_ZONE == 'AST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'AST',]$BEGIN_TIME, tzone = 'America/Halifax', roll = TRUE)
StormData[StormData$TIME_ZONE == 'CST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'CST',]$BEGIN_TIME, tzone = 'America/Chicago', roll = TRUE)
StormData[StormData$TIME_ZONE == 'EST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'EST',]$BEGIN_TIME, tzone = 'America/New_York', roll = TRUE)
StormData[StormData$TIME_ZONE == 'HST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'HST',]$BEGIN_TIME, tzone = 'Pacific/Honolulu', roll = TRUE)
StormData[StormData$TIME_ZONE == 'MST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'MST',]$BEGIN_TIME, tzone = 'America/Denver', roll = TRUE)
StormData[StormData$TIME_ZONE == 'PST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'PST',]$BEGIN_TIME, tzone = 'America/Los_Angeles', roll = TRUE)
StormData[StormData$TIME_ZONE == 'SST',]$BEGIN_TIME <- 
  force_tz(StormData[StormData$TIME_ZONE == 'SST',]$BEGIN_TIME, tzone = 'Pacific/Samoa', roll = TRUE)
### other timezones 'GMT', 'GST', and 'UTC' all appear to be UTC+0, and so left alone

```

### Event synonyms

Some events in **EVTYPE** which appear to be the same are recorded with different names (synonyms)

e.g. 'TSTM WIND', 'THUNDERSTORM WIND', 'THUNDERSTORM WINDS'

These synonyms should be merged.

In this section 'close' synonyms will not be merged e.g. 'MARINE THUNDERSTORM WIND' will not be merged with 'THUNDERSTORM WIND', 'THUNDERSTORM WIND/HAIL' will not be merged into 'THUNDERSTORM WIND' and 'STRONG WIND' will
not be merged into 'WIND', although arguably some of these close synonyms are the same.


``` {r}

### Most revents are recorded in capitals ('upper case'), but some are recorded in mixed (upper and lower) case
### Covert all to upper case, strip all numbers, periods and trailing/leading whitespace
### numbers and parentheses are often used to describe sub-types within an event type e.g. size of hail
### there are some subtypes which are not yet merged with this logic e.g. 'G45' sub-types
StormData$EVTYPE <- as.factor(toupper(trimws(gsub('[0-9,.,(,)]+','',StormData$EVTYPE))))

### list of synonyms
synonyms <- list(c('THUNDERSTORM WIND','TSTM WIND','THUNDERSTORM WINDS','THUNDERSTORM WINDSS','THUNDERSTORMS WINDS','THUNDERSTROM WIND','THUNDERTSORM WIND','THUNERSTORM WINDS','TUNDERSTORM WIND','THUNDERSTORM WINS','TSTM WND'),
                 c('MARINE THUNDERSTORM WIND','MARINE TSTM WIND','MARINE TSTM WINDS'),
                 c('HIGH WIND','STRONG WIND','HIGH WINDS','STRONG WINDS'),
                 c('FLASH FLOOD','FLASH FLOODING','FLASH FLOODS'),
                 c('FLOOD','FLOODING'),
                 c('URBAN FLOOD','URBAN FLOODING'),
                 c('WILDFIRE','WILD/FOREST FIRE'),
                 c('RECORD HEAT','RECORD WARMTH'),
                 c('THUNDERSTORM WINDS HAIL','TSTM WIND/HAIL','THUNDERSTORM WINDS/HAIL'),
                 c('WIND GUSTS','GUSTY WIND'),
                 c('TORRENTIAL RAIN','TORRENTIAL RAINFALL'),
                 c('TORNADO','TORNADOS','TORNDAO'),
                 c('WIND','WND'),
                 c('THUNDERSTORM','THUNDERSTORMS'),
                 c('RIP CURRENT', 'RIP CURRENTS'),
                 c('HURRICANE','TYPHOON','HURRICANE/TYPHOON'))

### merge synonyms into one synonym
for (events in synonyms) {
  StormData[as.character(StormData$EVTYPE) %in% events,]$EVTYPE <- events[1]
}
```

### Calculating economic damage

First need to calculate 'true' dollar values of economic damage, combining
**PROPDMG** with **PROPDMGEXP** and **CROPDMG** and **CROPDMGEXP**

In [National Weather Service](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
there are only three recognized abbreviations for **PROPDMGEXP** and **CROPDMGEXP**, being 'K' for thousands,
'M' for millions and 'B' for billions.

``` {r}
### convert the property and crop damage variables to true numeric values
StormData <- StormData %>%
  mutate(PropertyValue = PROPDMG * 1000^match(toupper(PROPDMGEXP),c('K','M','B'), nomatch=0)) %>%
  mutate(CropValue = CROPDMG * 1000^match(toupper(CROPDMGEXP),c('K','M','B'), nomatch=0))

```

### Finding data which appears to have been collected consistently with the most recent years

Data has been collected from 1950 to 2011. Has data been consistently collected during this time?
Compare the change in events recorded annually from 1950 to 2011, including the total number of
events and the five most commonly recorded events.

``` {r}
### find the most common event types
frequent.events <- StormData %>%
  count(EVTYPE) %>%
  arrange(desc(n))

# show the twenty most frequent event types
print(head(frequent.events,20))
```

Plot the yearly event count for the top ten events,
and the total event count (the total event count includes both 'common' and 'uncommon' events)

``` {r}

# the ten most frequent events
most.frequent.events <- as.list(as.character(frequent.events[1:10,]$EVTYPE))

# count the events in 'most.frequent.events' in each year of recording
frequent.event.count <- StormData %>%
  filter(EVTYPE %in% most.frequent.events) %>%
  mutate(event.year = year(BEGIN_TIME)) %>%
  group_by(event.year, EVTYPE) %>%
  summarise(events = n())

# and the total events for each year
yearly.event.count <- StormData %>%
  mutate(event.year = year(BEGIN_TIME)) %>%
  group_by(event.year) %>%
  summarise(events = n())

yearly.event.count[,'EVTYPE'] <- 'TOTAL'

event.count <- bind_rows(frequent.event.count,yearly.event.count)

# 'normalize the events' to the frequency observed in 2011
# store the normalized event count in 'event.norm'
event.count[,'event.norm'] <- NA

for (event in list.append(most.frequent.events,'TOTAL')) {
  event.2011.count <- with(event.count, events[EVTYPE == event & event.year == 2011])
  for (year in with(event.count, event.year[EVTYPE == event])) {
    index <- with(event.count,(EVTYPE == event) & (event.year == year))
    event.count$event.norm[index] <- (event.count$events[index]/event.2011.count)
  }
}

ggplot(data=event.count,aes(x=event.year,y=event.norm,group=EVTYPE, colour=EVTYPE)) +
  geom_line()+geom_point() +
  labs(title = "Ten most frequent storm events by year, and total events, from 1950 to 2011",
       subtitle = "NOAA storm database. Yearly event frequency in comparison to 2011 event frequency") +
  labs(x = "Year", y = "Frequency in comparison to 2011 frequency") +
  theme(legend.position=c(0.2,0.7))


```

Many of the frequent events are not recorded in the current form until at the mid-1990s.
'MARINE THUNEDERSTORM WIND' is not even recorded until 2001.

Restrict data analysis from 1997 to 2011 (fifteen years in total)

``` {r}

StormSubset <- StormData[year(StormData$BEGIN_TIME)>=1997,
                         c('COUNTYNAME','STATE','EVTYPE','FATALITIES','INJURIES',
                           'PROPDMG','PROPDMGEXP','CROPDMG','CROPDMGEXP',
                           'REFNUM','BEGIN_TIME')]



```

## Which types of events are most harmful with respect to population health?

Rank the twenty events with the greatest number of fatalities or injuries per year.
Population health harm contained in variables **FATALITIES** and **INJURIES**.

``` {r}

### calculate the sum of fatalities and injuries for each event for each year
HealthSum <- StormSubset %>%
  mutate(event.year = year(BEGIN_TIME)) %>%
  group_by(event.year, EVTYPE) %>%
  summarise(fatalities = sum(FATALITIES), injuries = sum(INJURIES))

### calculate the mean ('average') annual fatalities and injuries for each event
### calculate the proportion of annual fatalities or injuries caused by the event 
HealthMean <- HealthSum %>%
  group_by(EVTYPE) %>%
  summarise(mean_fatalities = mean(fatalities), mean_injuries = mean(injuries)) %>%
  mutate(proportion_fatalities = mean_fatalities/sum(mean_fatalities)*100) %>%
  mutate(proportion_injuries = mean_injuries/sum(mean_injuries)*100) %>%
  arrange(desc(mean_fatalities))

kable(HealthMean[1:20,c('EVTYPE','mean_fatalities','proportion_fatalities')], caption = 'Weather events causing most fatalities, annual mean fatalities, and percentage of fatalities caused by event')

total_mean_fatalities <- sum(HealthMean$mean_fatalities)
sprintf('Total mean number of fatalities annually: %.1f\n',total_mean_fatalities)

kable(arrange(HealthMean,desc(mean_injuries))[1:20,c('EVTYPE','mean_injuries','proportion_injuries')],caption = 'Weather events causing the most injuries, annual mean injuries and percentage of injuries caused by event')

total_mean_injuries <- sum(HealthMean$mean_injuries)
sprintf('Total mean number of injuries annually: %.1f',total_mean_injuries)

```

Display results graphically of the events causing the most fatalities and injuries

``` {r}

### the (union) of the ten events causing the most fatalities and ten events causing the most injuries
most_health_events <- union(HealthMean$EVTYPE[1:10],arrange(HealthMean,desc(mean_injuries))$EVTYPE[1:10])

### gather data in preparation for bar-charting
HealthImpact <- HealthMean %>%
  filter(HealthMean$EVTYPE %in% most_health_events) %>%
  mutate(Injuries = -proportion_injuries) %>% ### negative values will be used for 'bar-negative-stack' later
  rename(Fatalities = proportion_fatalities) %>%
  select(one_of('EVTYPE','Injuries','Fatalities')) %>%
  gather(key = 'harm', value = 'proportion', Injuries, Fatalities)

subtitle <- sprintf('Data from NOAA storm database 1997 to 2011\n
                    Total mean fatalities per year : %.1f\n
                    Total mean injuries per year : %.1f',
                    total_mean_fatalities, total_mean_injuries)  

ggplot(data=HealthImpact, aes(x = EVTYPE, y = proportion, fill = harm)) +
  geom_bar(stat = 'identity') + ### 'identity' needed for values (rather than counts)
  labs(title = "The storm events causing the highest proportions of fatalities and injuries",
       subtitle = subtitle) +
  labs(x = "Event", y = "Percentage proportion of annual fatalities or injuries") +
  scale_y_continuous(labels=abs) + 
                     ### remove negative values for axis for 'negative' part of chart
  guides(fill = guide_legend(title="")) +
  theme(legend.position=c(0.8,0.5)) +
  coord_flip(ylim=c(-35,35))

```

## Which types of events have the greatest economic consequences?

Rank the twenty events with the greatest value of property or crop damage per year
Economic damage information is stored in **PropertyValue** and **CropValue**.

``` {r}
### remove variables no longer required
DamageSubset <- StormSubset %>%
  select(-c(PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,FATALITIES,INJURIES))

### calculate the sum of damage for each event for each year
DamageSum <- DamageSubset %>%
  mutate(event.year = year(BEGIN_TIME)) %>%
  group_by(event.year, EVTYPE) %>%
  summarise(property = sum(PropertyValue), crop = sum(CropValue))

### calculate the mean ('average') damage for each event (for each of property and crop)
### calculate the proportion of annual damage caused by event (for each of property and crop)
### and also calculate combined mean damage (property+crop) and proportion
DamageMean <- DamageSum %>%
  group_by(EVTYPE) %>%
  summarise(mean_property = mean(property), mean_crop = mean(crop)) %>%
  mutate(mean_combined = mean_property + mean_crop) %>%
  mutate(proportion_property = mean_property/sum(mean_property)*100) %>%
  mutate(proportion_crop = mean_crop/sum(mean_crop)*100) %>%
  mutate(proportion_combined = mean_combined/sum(mean_combined)*100) %>%
  arrange(desc(mean_property))

kable(DamageMean[1:20,c('EVTYPE','mean_property','proportion_property')], caption = 'Weather events causing most property damage, annual mean proeprty damage, and percentage of property damage caused by event')

total_mean_property <- sum(DamageMean$mean_property)
sprintf('Total mean value of property damage annually: $%.0f\n',total_mean_property)

kable(arrange(DamageMean,desc(mean_crop))[1:20,c('EVTYPE','mean_crop','proportion_crop')],caption = 'Weather events causing the most crop damage, annual mean crop damage and percentage of crop damage caused by event')

total_mean_crop <- sum(DamageMean$mean_crop)
sprintf('Total mean value of crop damage annually: $%.0f\n',total_mean_crop)

kable(arrange(DamageMean,desc(mean_combined))[1:20,c('EVTYPE','mean_combined','proportion_combined')],caption = 'Weather events causing the most economic damage, annual mean economic damage and percentage of economic damage caused by event')

total_mean_combined <- sum(DamageMean$mean_combined)
sprintf('Total mean economic damage annually: $%.0f',total_mean_combined)

```

Display results graphically of the events causing the most property and crop damage

``` {r}

### the (union) of the ten events causing the most property, crop or combined economic damage
most_damage_events <- Reduce(union, list(DamageMean$EVTYPE[1:10],arrange(DamageMean,desc(mean_combined))$EVTYPE[1:10],
                            arrange(DamageMean,desc(mean_combined))$EVTYPE[1:10]))

### gather data in preparation for bar-charting
DamageImpact <- DamageMean %>%
  filter(DamageMean$EVTYPE %in% most_damage_events) %>%
  rename(Property = proportion_property, Crop = proportion_crop, Combined = proportion_combined) %>%
  select(one_of('EVTYPE','Property','Crop','Combined')) %>%
  gather(key = 'damage', value = 'proportion', Property, Crop, Combined)

subtitle <- sprintf('Data from NOAA storm database 1997 to 2011\n
                    Total mean property damage per year \t: $%12.0f\n
                    Total mean crop damage per year \t\t: $%12.0f\n
                    Total combined damage per year \t\t: $%12.0f\n',
                    total_mean_property, total_mean_crop, total_mean_combined)  

ggplot(data=DamageImpact, aes(x = EVTYPE, y = proportion, fill = damage)) +
  geom_bar(stat = 'identity', position = 'dodge') + ### 'identity' needed for values (rather than counts)
  ### 'dodge' for side-by-side bar chart
  labs(title = "The storm events causing the highest proportions of economic impact",
       subtitle = subtitle) +
  labs(x = "Event", y = "Percentage proportion of property, crop or combined value of loss") +
  guides(fill = guide_legend(title="")) +
  theme(legend.position=c(0.8,0.5)) +
  coord_flip()

```
